---
title: "EDA: numeric"
author: "Longyi Zhao"
date: "2023-10-05"
output: github_document 
---

```{r}
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10, 
    month = lubridate::floor_date(date, unit = "month")) |>
  select(name, id, everything())
```
lubridate:floor_date: round to month, for example 2021-07-07 to 2021-07-01

## Initial numeric work

```{r}
weather_df |>
  ggplot(aes(x=prcp))+
  geom_histogram()
```

here are the big outliers

```{r}
weather_df |>
  filter(prcp>1000)
```
can double check to see if they are real data. 

```{r}
weather_df |>
  filter(tmax>=20, tmax <=30) |>
  ggplot(aes(x=tmin, y=tmax, color = name))+
  geom_point()
```
report F then convert to C, 

## Grouping

```{r}
weather_df |>
  group_by(name, month)
```
not human readable 

## counting

```{r}
weather_df |>
  group_by(name, month)|>
  summarize(n_obs = n())
```
for central park in january 2021, 31 observations

```{r}
weather_df |>
  group_by(month)|>
  summarize(n_obs = n())
```
number of people in the treatment 

```{r}
weather_df|>
  count(name, name = "n_obs")
```

```{r}
weather_df |>
  count(name, month) |>
  pivot_wider(
    names_from = name, 
    values_from = n
  )
```

## general summarize

```{r}
weather_df |>
  drop_na(tmax) |>
  group_by (name) |>
  summarize(
    mean_tmax = mean(tmax), 
    median_tmax = median(tmax, na.rm = TRUE), 
    sd_tmax = sd(tmax, na.rm = TRUE)
  ) 

```

```{r}
weather_df |>
  group_by (name, month)|>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) |>
  ggplot(aes(x=month, y=mean_tmax, color = name)) +
  geom_point()+
  geom_line()
```

```{r}
weather_df |>
  group_by (name, month)|>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) |>
  pivot_wider(
    names_from = name, 
    values_from = mean_tmax
  ) |>
  knitr::kable(digits = 2)
```
knitrlabel: create a table to help understand plot

## Grouped Mutate

```{r}
weather_df |>
  group_by (name) |>
  mutate(mean_tmax = mean(tmax, na.rm = TRUE))
```

```{r}
weather_df |>
  group_by (name) |>
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE), 
    centerd_tmax = tmax-mean_tmax) |>
  ggplot(aes(x = date, y = centerd_tmax, color = name)) +
  geom_point()
```

```{r}
weather_df |>
  group_by (name, month)|>
  mutate(tmax_rank = min_rank(desc(tmax)))|>
  filter(tmax_rank < 2)
```
desc: from highest to lowest

lags
```{r}
weather_df |>
  group_by (name)|>
  mutate(yesterday_tmax = lag(tmax))
```
showing yesterday's tmax in another column 
mutate(yesterday_tmax = lag(tmax, 3)) temperature three days back 

```{r}
weather_df |>
  group_by (name) |>
  mutate(temp_change = tmax - lag(tmax)) |>
  summarize(
    sd_temp_change = sd(temp_change, na.rm = TRUE)
  )
```
day to day variation of temperature at these locations 

















